Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-04-27 paper/manuscript.Rmd
tweaks to paper

PSYCH 251 is Stanford Psychology's graduate-level experimental methods class taught by MCF.  During the 10 week class, students replicate a published finding. They individually re-implement the study, write analysis code, pre-register their study, collect data using an online platform, and write up a structured replication report. Students are free to choose studies related to their research interests, with the default recommendation being an article from a recent year of Psychological Science. While this choice results in a non-random sample from the literature, the sample is representative of studies that are of interest to and doable by first year graduate students.

The sample of replicated studies reflects the variability of the literature, including studies from different subfields, using different experimental methods and statistical outcomes. We leverage the naturally occurring variability in this sample of replications to examine how different demographic, experimental design, and statistical properties predict replication success.

Many different measures can be used to define replication success of an individual statistical result, and there is much discussion over what each measure represents and which make sense [@simonsohn2015; @gelman2018; @mathur2020]. Because we operationalized replicability as whether a study could be built upon, we used a subjective rating of replication success as our primary outcome measure. This measure has the benefit that it was applicable across the diverse range of statistical measures and reporting practices present in the sample. It, unlike statistical measures of replication, could easily accommodate studies where there were multiple important outcome measures that together defined the pattern of interest. A holistic measure of replication success had been coded for each project when it was turned in at the end of the class. For reliability, VB independently code the replication success from the replication reports; discrepancies were resolved by discussion between MCF and VB (`r round(pluck(matching,2,1)/176*100)`% of cases).

As a complement, we also used two statistical measure of replication on the subset of the data where they were computable (`r pluck(pred_int_avail,1)` cases, see Figure \ref{fig:prisma}). We measured p-original, the p-value on the null hypothesis that the original and replication statistics are from the same distribution, as a continuous variable, and we also determined whether the replication statistic fell within the prediction interval of the original statistic [@errington2021]. These both measure how similar the estimates from the two sets of data are.

@@

pred_int_rate <- d |> filter(!is.na(predInt)) |> summarize(m=mean(as.numeric(predInt))*100)
porig_rate <- d |> filter(!is.na(p_orig)) |> summarize(m=median(p_orig))
```

Across the 176 studies, the overall subjective replication rate was `r round(pluck(sub_rep_rate,1))`%. `r round(pluck(pred_int_rate,1))`% (N=FOOBAR) of the studies had replication outcomes within the prediction interval of the original outcome. The median p_original value was `r round(pluck(porig_rate, 1),2)`. Figure \ref{fig:smd} shows the relationship between original SMD, replication SMD, and subjective replication score. Roughly speaking, there's a cluster of studies that replicate with similar effect sizes to the original and another cluster that fail to replicate with replication effect sizes near zero. On average, there is a diminution of effect sizes from original to replication.

@@

## Single predictors
Properties of both the original study and the replication can influence whether or not the replication is a success. We chose a set of predictor variables from the correlational results of RP:P and our own intuitions about experimental factors that might impact replication success as well as some covariates related to how close the replication would be. A full description of these features is given in methods.

Many of these predictors individually correlate with subjective replication success (Table \ref{tab:cor}). Predictors of higher replicability included within-subjects designs, higher numbers of trials, and open data. Predictors of lower replicability included single vignetted studies, social psychology studies, and original-replication pairs where the replication switched to online.

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented, and the cognitive psychology studies replicated at `r round(1/subfield, 2)` times the rate of social psychology studies. Within and between subjects designs were both common, and within replicated `r within |> round(2)` times as much. Similarly, studies with multiple vignettes replicated `r round(1/vignette, 2)` times more than single vignetted studies. However, there were strong correlations among these experimental features and between these experimental features and subfield.

Studies with open data, which almost always also had open materials, tended to replicate more than studies without open data, although this may be linked to temporal trends.

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person).  While online studies in general show comparable results to studies conducted in person [@crump2013], switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

## Regression model
While a number of the predictors show individual correlations with the subjective replication score, many of the predictors are also correlated with one another. In order to determine which predictors were the strongest, we ran a series of pre-registered regularized regression models (see Methods for details; see Supplement for all estimates from all models). The coefficient estimates of the primary model, predicting the subjective replication scores based on all the data, are shown in Figure \ref{fig:mod-results}. Due to a large number of predictors coupled with a small and noisy dataset, even with strong regularization, there is much uncertainty around the coefficients. The general directions of coefficients are consistent with the effects of the predictors in isolation.

@@

We also ran models predicting whether the replication effect was within the prediction interval as the original effect and what the p-original was between the replication and original. Both these models had even more uncertain estimates. While the credible intervals are wide, the general patterns of predictors are similar to the subjective replication models. The strongest predictors are still within-subjects designs (`r within_2`) and studies with larger effect sizes (`r smd_2`).

@@

# Discussion

The replication rate in psychology is, to the best estimates, somewhere around 50%, which is quite low. Many reasons for this low replication rate have been hypothesized, but studying it either experimentally or correlationally requires doing a number of replication experiments, which can be a lot of work. Here we took advantage of 11 years of graduate student replication projects to look at correlational predictors of replication in a previously-unused dataset.

In line with previous replications, we found a FOOBAR replication rate, with some studies showing effect sizes similar to the original and others much smaller. Some individual correlates of replicability that stood out included within-subjects designs, work in the subfield of cognitive psychology, and the original and replication both using online samples.  Many of these correlates interrelate with one another, and we are still limited in our sample, so models with multiple predictors have a lot of uncertainty around the effects of each predictor.

We are explicit in our goal of estimating how likely it is that a first-year graduate student, attempting to replicate a study of interest to them, gets replication results which are consistent enough with the original results that the student could build on the study in their own work. We can't estimate some platonic probability of truth for a study and neither can any replication study. Instead, we can estimate how likely studies are to get results of different levels of closeness in different circumstances. Much of the discussion around whether things should count as replication successes, or failures, or not replications at all is really about what the right thresholds are for closeness of results, and what the right circumstances are for a replication. Rather than try to find the one right answer to whether a direct replication was done in a certain right way, we think it's useful to explicitly label what the estimand really was: how likely it is to get the same results (under whatever metric) given circumstances like those used for the replication.

In our case, our estimand is about how well first-year graduate students will do at the replication, which takes into account the limited time, limited budget, and limited experience. We think this is an important estimand, as much of the work of psychology is done by trainees, in circumstances like these. Thus, if some studies have delicate results that require large samples and very exact methods to achieve, they may not replicate under normal resources.

@@

```
The dataset of replication projects comes from class projects conducted in PSYCH 251 (earlier called PSYCH 254) a graduate-level experimental methods class taught by MCF from 2011 to 2022. This class is commonly taken by first year graduate students in psychology and related disciplines, and it has been a requirement of the Psychology PhD since around 2015. Each student chose a study to replicate, implemented the study, wrote analysis code, pre-registered their replication, ran the study, and turned in a structured final report including methods, analytic plan, changes from the original study, confirmatory and exploratory analyses, and discussion of outcomes. Students were encouraged to do experimental replications, but some students chose to replicate correlational outcomes or do computational reproducibility projects instead. We cannot include the full student reports for confidentiality reasons, but we include an example as well as the template given to students in the repo. TODO example and template

Students were free to choose what study they replicated; the recommended path for students who did not have their own ideas was to pick an interesting study from a recent year of Psychological Science (this led to a high fraction of Psych Science articles in the replication sample, `r nrow(ps)`, `r nrow(ps)/nrow(all)*100 |> round()`% of studies).

@@

### Measures of replication success
Our primary replication outcome is experimenter and instructor rated replication success (0-1). The subjective replication success was recorded by the teaching staff for the majority of class replications at the time they were conducted. Where the values were missing they were filled in by MCF on the basis of the reports. For all studies, replication success was independently coded by VB on the basis of the reports. Where VB’s coding disagreed with the staff/MCF’s code, the difference was resolved by discussion between VB and MCF (`r pluck(matching,2,1)/176*100 |> round()`% of studies). These were coded on a [0, .25, .5, .75, 1] scale.

@@

We coded the number of subjects, post-exclusions. We coded whether a study had a between-subjects, within-subjects, or mixed design; for analyses mixed studies were counted as within-subjects designs. In the analysis, we used a log-scale for number of subjects and numbers of trials.

### Properties of replication
We coded whether the replication was conducted on a crowd-sourced platform; this was the norm for the class projects, but a few were done in person. As the predictor variable, we used whether the recruitment platform was changed between original and replication. This groups the few in-person replications in with the studies that were originally online and stayed online in a "no change" condition, in contrast with the studies that were originally in-person with online replications.

@@

As a control variable, we included whether the original authors were faculty at Stanford at the time of the replication. This is to account for potential non-independence of the replication (ex. if replicating their advisor's work, students may have access to extra information about methods).

We made note of studies to exclude from some of the sensitivity analyses, due to not quite aligned statistics, extremely small or unbalanced sample sizes, or where the key statistical measure the student chose was not of central importance to the original study.

@@

In some cases, p-values and or effect sizes were not calculable either due to insufficient reporting (ex. reporting a p-value but no other statistics from a test) or key measures where p-values and effect sizes did not apply (ex. PCA as measure of interest). Where studies reported beta estimates and standard errors or proportions, SMD isn't an applicable measure, but we were still able to calculate p_original and prediction interval.

We separately coded whether the original and replication effects were in the same direction, using raw means and graphs. This is more reliable than the statistics because F-tests don't include the direction of effect, and some students may have flipped the direction in coding for betas or t-tests. In the processed data, the direction of the effect of the replication was always coded consistently with the original study’s coding, so a positive effect was in the same direction as the original and a negative effect in the opposite direction.

In regressions, we used SMD and log p-value as predictors.

@@

Due to the monotonic missingness of the data, we had more predictor variables and outcome variables for some original-replication pairs than others. To take full advantage of the data, we ran a series of models, with some models having fewer predictors, but more data, and others having more predictors, but more limited data.

We ran a model predicting the subjective replication score on the basis of demographic and experimental predictors on the entire dataset; we ran two models predicting p_original and prediction interval from demographic and experimental predictors on the subset of data where we had p_original and prediction intervals. Then, on the smaller subset of the data where we had SMD and p-values, we re-ran these three models with those as additional predictor variables.

>>>>

PSYCH 251 is Stanford Psychology's graduate-level experimental methods class taught by MCF.  During the 10 week class, students replicated a published finding. They individually re-implemented the study, wrote analysis code, pre-registered their study, collected data using an online platform, and wrote up a structured replication report. Students were free to choose studies related to their research interests, with the default recommendation being an article from a recent year of Psychological Science. The resultant sample of studies is not a random sample from the literature but is representative of studies that are of interest to and doable by first year graduate students.

The sample of replicated studies reflect the variability of the literature, including studies from different subfields, using different experimental methods and statistical outcomes. We leveraged the naturally occurring variability in this sample of replications to examine how different demographic, experimental design, and statistical properties predict replication success.

Many different measures can be used to define replication success of an individual statistical result [@simonsohn2015; @gelman2018; @mathur2020]. Because we operationalized replicability as whether a study could be built upon, we used a subjective rating of replication success as our primary outcome measure. This measure was applicable across the diverse range of statistical measures and reporting practices present in the sample. It, unlike statistical measures of replication, could easily accommodate studies where there were multiple important outcome measures that together defined the pattern of interest. A holistic measure of replication success had been coded for each project when it was turned in at the end of the class. For reliability, VB independently code the replication success from the replication reports; discrepancies were resolved by discussion between MCF and VB (`r round(pluck(matching,2,1)/176*100)`% of cases).

As a complement, we also used two statistical measure of replication on the subset of the data where they were computable (`r pluck(pred_int_avail,1)` cases, see Figure \ref{fig:prisma}). We measured p-original, the p-value on the null hypothesis that the original and replication statistics are from the same distribution, as a continuous variable, and we also determined whether the replication statistic fell within the prediction interval of the original statistic [@errington2021]. These both measure how similar the outcome estimates from the two sets of data are.

@@

pred_int_rate <- d |> filter(!is.na(predInt)) |> summarize(m=mean(as.numeric(predInt))*100)
porig_rate <- d |> filter(!is.na(p_orig)) |> summarize(m=median(p_orig))

d |> filter(!is.na(predInt)) |> filter(predInt) |> nrow() -> predIntcount
d |> filter(!is.na(predInt)) |> nrow() -> predInttotal

Across the 176 studies, the overall subjective replication rate was `r round(pluck(sub_rep_rate,1))`%. `r round(pluck(pred_int_rate,1))`% (`r predIntcount`/`r predInttotal`) of the studies had replication outcomes within the prediction interval of the original outcome. The median p_original value was `r round(pluck(porig_rate, 1),2)`. Figure \ref{fig:smd} shows the relationship between original standardized effect size, replication effect size, and subjective replication score. Roughly speaking, there's a cluster of studies that replicate with similar effect sizes to the original and another cluster that fail to replicate with replication effect sizes near zero. On average, there is a diminution of effect sizes from original to replication.

@@

## Single predictors
Properties of both the original study and the replication can influence whether or not the replication is a success. We chose a set of predictor variables from the correlational results of RP:P and our own intuitions about experimental factors that might impact replication success as well as some covariates related to how close the replication was. A full description of these features is given in Methods.

Many predictors individually correlate with subjective replication success (Table \ref{tab:cor}). Predictors of higher replicability included within-subjects designs, higher numbers of trials, and the original study having open data. Predictors of lower replicability included single vignetted studies, social psychology studies, and original-replication pairs where the original study was in-person and replication switched to online.

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented, and the cognitive psychology studies replicated at `r round(1/subfield, 2)` times the rate of social psychology studies. Within and between subjects designs were both common, and within-subjects designs replicated `r within |> round(2)` times as much. Similarly, studies with multiple vignettes replicated `r round(1/vignette, 2)` times more than single vignetted studies. However, there were strong correlations among these experimental features and between these experimental features and subfield. Studies with open data, which almost always also had open materials, tended to replicate more than studies without open data, although this may be linked to temporal trends.

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person).  While online studies in general show comparable results to studies conducted in person [@crump2013], switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well-adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

## Regression model
While a number of predictors show individual correlations with the subjective replication score, many of the predictors also correlate with one another. In order to determine which predictors were the strongest, we ran a series of pre-registered regularized regression models (see Methods for details; see Supplement for all estimates from all models). We ran models both using all the data, but without statistical predictors (that were uncodable for some studies) and models including statistical predictors, but limiting to the subset of data where all predictors were available. The coefficient estimates from two models predicting the subjective replication scores are shown in Figure \ref{fig:mod-results}. Due to a large number of predictors coupled with a small and noisy dataset, even with strong regularization, there is much uncertainty around the coefficients. The general directions of coefficients are consistent with the effects of the predictors in isolation.

@@

We also ran models predicting our secondary outcome measures: whether the replication effect was within the prediction interval as the original effect and what the p-original was between the replication and original. Both these models had even more uncertain estimates. While the credible intervals were wide, the general patterns of predictor direction and relative strength were similar to the subjective replication models. The strongest predictors were still within-subjects designs (`r within_2`) and studies with larger effect sizes (`r smd_2`).

@@

# Discussion
Prior large scale replications of psychology findings have estimated the replication rate at somewhere around 50%. This rate feels low, raising the question of what is causing so many studies not to replicate. However, limited numbers of replications and limited research into specific predictors of replication failure leave this an open question.

Here we took advantage of 11 years of graduate student replication projects to look at correlational predictors of replication in a previously-unused dataset.

In line with previous results, we found a ` rround(pluck(sub_rep_rate,1))`% replication rate, with some studies showing effect sizes similar to the original and others much smaller. When we looked at individual correlates of replicability, within-subjects designs, work in the subfield of cognitive psychology, and the original and replication both using online samples stood out as the strongest correlates. As many of these predictors interrelate with one another, we ran regularized regressions with all the predictors at once. Due to our small sample, model estimates were uncertain, but within-subjects designs and large original effect sizes were the strongest predictors.

<!--We are explicit in our goal of estimating how likely it is that a first-year graduate student, attempting to replicate a study of interest to them, gets replication results which are consistent enough with the original results that the student could build on the study in their own work. We can't estimate some platonic probability of truth for a study and neither can any replication study. Instead, we can estimate how likely studies are to get results of different levels of closeness in different circumstances. Much of the discussion around whether things should count as replication successes, or failures, or not replications at all is really about what the right thresholds are for closeness of results, and what the right circumstances are for a replication. Rather than try to find the one right answer to whether a direct replication was done in a certain right way, we think it's useful to explicitly label what the estimand really was: how likely it is to get the same results (under whatever metric) given circumstances like those used for the replication. 
-->
In our case, our estimand is about how well first-year graduate students will do at the replication, which takes into account the limited time, limited budget, and limited experience. We think this is an important estimand, as much of the work of psychology is done by trainees, in circumstances like these. Thus, if some studies have delicate results that require large samples and very exact methods to achieve, they may not replicate under normal resources.

@@

```
The dataset of replication projects comes from class projects conducted in PSYCH 251 (earlier called PSYCH 254) a graduate-level experimental methods class taught at Stanford by MCF from 2011 to 2022. This class is commonly taken by first year graduate students in psychology and related disciplines, and it has been a requirement of the Psychology PhD since around 2015. Each student chose a study to replicate, implemented the study, wrote analysis code, pre-registered their replication, ran the study, and turned in a structured final report including methods, analytic plan, changes from the original study, confirmatory and exploratory analyses, and discussion of outcomes. Students were encouraged to do experimental replications, but some students chose to replicate correlational outcomes or do computational reproducibility projects instead. We cannot include the full student reports for confidentiality reasons, but we include an example as well as the template given to students at TODO example and template.

Students were free to choose what study to replicate; the recommended path for students who did not have their own ideas was to pick an interesting study from a recent year of Psychological Science (this led to a high fraction of Psych Science articles in the replication sample, `r nrow(ps)`, `r nrow(ps)/nrow(all)*100 |> round()`% of studies).

@@

### Measures of replication success
Our primary replication outcome was experimenter and instructor rated replication success. The subjective replication success was recorded by the teaching staff for the majority of class replications at the time they were conducted. Where the values were missing they were filled in by MCF on the basis of the reports. For all studies, replication success was independently coded by VB on the basis of the reports. Where VB’s coding disagreed with the staff/MCF’s code, the difference was resolved by discussion between VB and MCF (`r pluck(matching,2,1)/176*100 |> round()`% of studies). Subjective replication scores were coded on a [0, .25, .5, .75, 1] scale.

@@

We coded the number of subjects, post-exclusions. We coded whether a study had a between-subjects, within-subjects, or mixed design; for the analysis, mixed studies were counted as within-subjects designs. In the analysis, we used a log-scale for number of subjects and numbers of trials.

### Properties of replication
We coded whether the replication was conducted on a crowd-sourced platform; this was the norm for the class projects, but a few were done in-person. For analysis, we coded this into a variable indicating if the recruitment platform changed between original and replication. This grouped the few in-person replications in with the studies that were originally online and stayed online in a "no change" condition, in contrast with the studies that were originally in-person with online replications.

@@

As a control variable, we included whether the original authors were faculty at Stanford at the time of the replication. This was to account for potential non-independence of these replications (ex. if replicating their advisor's work, students may have access to extra information about methods).

We made note of studies to exclude for sensitivity analyses, due to not quite aligned statistics, extremely small or unbalanced sample sizes, or a student choosing a key statistical measure that was not of central importance to the original study.

@@

In some cases, p-values and or effect sizes were not calculable either due to insufficient reporting (ex. reporting a p-value but no other statistics from a test) or key measures where p-values and effect sizes did not apply (ex. PCA as measure of interest). Where studies reported beta estimates and standard errors or proportions, standardized effects sizes are not an applicable measure, but we were still able to calculate p_original and prediction interval.

We separately coded whether the original and replication effects were in the same direction, based on raw means and graphs. This is more reliable than the statistics because F-tests don't include the direction of effect, and some students may have flipped the direction in coding for betas or t-tests. In the processed data, the direction of the effect of the replication was always coded consistently with the original study’s coding, so a positive effect was in the same direction as the original and a negative effect in the opposite direction.

In regression analyses, we used SMD and log p-value as predictors.

@@

Due to the monotonic missingness of the data, we had more predictor variables and outcome variables for some original-replication pairs than for others. To take full advantage of the data, we ran a series of models, with some models having fewer predictors, but more data, and others having more predictors, but less data.

We ran a model predicting the subjective replication score on the basis of demographic and experimental predictors on the entire dataset. We ran two models predicting p_original and prediction interval from demographic and experimental predictors on the subset of data where we had p_original and prediction intervals. Then, on the smaller subset of the data where we had effect sizes and p-values, we re-ran these three models with those as additional predictor variables.
